---
title: "Master script for postfire analysis"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### 1. Source functions, get data and plot

First we'll _source()_ (i.e. "run all code in") the scripts with the functions we made. Then we'll set the URL, read in the data with _download.NDVI()_, and plot it with _plot.NDVI()_.

```{r}
## Load required functions by running source() on the individual function files
if(file.exists("01_download.NDVI.R")) source("01_download.NDVI.R")
if(file.exists("02_plot.NDVI.R"))     source("02_plot.NDVI.R")
if(file.exists("03_negexp.R"))        source("03_negexp.R")

## Download NDVI data
URL = "https://raw.githubusercontent.com/jslingsby/BIO3019S_Ecoforecasting/master/data/modisdata.csv"
dat <- download.NDVI(URL)

# Convert "calendar_date" to postfire age in days since fire - assuming the first date in the times eries is the time of the fire 
dat$age <- (as.numeric(dat$calendar_date) - min(as.numeric(dat$calendar_date), na.rm = T))/365.25

## Plot overall NDVI time series
plot.NDVI(dat)
```

<br>

Q1: This plot suggests that Fynbos greenness (NDVI) as observed from satellite saturates with time since fire. Why do you think it saturates rather than increasing linearly with time?

>*Answer 1:*
after the fire, the vegetation grows rapidly as there is open space for seedlings to germinate and grow. However, it saturates over time because as the vegetation covers all of the space of one pixel. Even if there is more growth, it can only be vertical so the greenness measure cannot pick up that growth since the greenness remains the same. 
<br>

### 2. Fit models using Non-linear Least Squares (NLS)

Now we'll fit the simple and full negative exponential models using Non-linear Least Squares (NLS).

First the simpler model:

```{r}
## Simple model

# set parameters
par <- c(alpha = 0.2, gamma = 0.4, lambda = 0.5)

# fit model
fit_negexp <- nls(NDVI ~ alpha + gamma * (1 - exp(- age/lambda)),
                  data = dat, start = par, trace = F, 
                  control = nls.control(maxiter = 500))

# plot
plot.NDVI(dat = dat, fit = fit_negexp)
```

<br>

And let's look at the model summary with parameter estimates

```{r}
# print model summary
summary(fit_negexp)
```

<br>

Now the full model:

```{r}
## Full model

# set parameters
par <- c(alpha = 0.2, gamma = 0.4, lambda = 0.5, A = 0.6, phi = 0)

# fit model
fit_negexpS <- nls(NDVI ~ alpha + gamma * (1 - exp(- age/lambda))
                   + A*sin(2*pi*age + (phi + pi/6*(3 - 1))), 
                   data = dat, start = par, trace = F, 
                   control = nls.control(maxiter = 500))

# plot
plot.NDVI(dat = dat, fit = fit_negexpS)

```


```{r}
# print model summary
summary(fit_negexpS)
```

<br>

Lots more parameters...

Q2: How do the estimates for the common parameters compare?

>*Answer 2:*
alpha and gamma are slightly lower from the first to the second model but lambda increases from the first to the second 
model.
<br>

### 3. Compare NLS models using ANOVA

Modelers often want to know which of a set of models are better. One way to do this when comparing nested* models using least squares is using analysis of variance (ANOVA). In this case the `anova()` function will take the model objects as arguments, and return an ANOVA testing whether the full model results in a significant reduction in the residual sum of squares (and thus is better at capturing the data), returning an F-statistic, Degrees of Freedom (the difference in the number of parameters between the models) and p-value.

*i.e. one model is a subset of the other, as in our case

```{r}
anova(fit_negexp, fit_negexpS)
```

<br>

Q3: Which model is better?

>*Answer 3:*
Model 2 is better since model 2 has a more minimised sum of squared residuals compared to model 1. This could be due to model 2 accounting for seasonality so it is a better fit to the data.
Q4: How many degrees of freedom are there in this ANOVA and why (i.e. what are they)?

>*Answer 4:*
2 because there are 2 more parameters in model 2 than in model 1 which are A and phi.
<br>

### 4. Fit models using Maximum Likelihood Estimation (MLE)

First let's fit the simpler model:

```{r}
## Fit the simpler model using MLE

# set parameters
par <- c(alpha = 0.2, gamma = 0.4, lambda = 0.5)

# fit model
fit_negexpMLE <- fit.negexp.MLE(dat, par)

# plot
plot.NDVI(dat)
# add curve with MLE parameters
lines(dat$age, pred.negexp(fit_negexpMLE$par,dat$age), col = 'skyblue', lwd = 3)

```


```{r}
fit_negexpMLE
```

<br>

Then the full model:

```{r}
## Fit the full model using MLE

# set parameters
par <- c(alpha = 0.2, gamma = 0.4, lambda = 0.5, A = 0.6, phi = 0)

# fit model
fit_negexpMLES <- fit.negexpS.MLE(dat, par)

# plot
plot.NDVI(dat)
# add curve with MLE parameters
lines(dat$age, pred.negexpS(fit_negexpMLES$par,dat$age), col = 'skyblue', lwd = 3)
```

```{r}
fit_negexpMLES
```

<br>

### 5. Compare MLE models using Akaike's information criterion (AIC)

Note that we can't compare our MLE models using ANOVA because our custom functions do not return full model fits like the `nls()` function - only the parameter estimates, negative log-likelihoods and a few other diagnostics.

Another way to compare models (and probably the most common) is using the Akaike information criterion (AIC), which is an estimator of prediction error (i.e. relative quality) of statistical models for a given set of data. 

The formula for the Akaike information criterion is:

$AIC = 2K -2(ln(L))$

Where:

- $k$ = the number of estimated parameters in the model
- $L$ = maximum value of the likelihood function for the model

Since we have our negative log likelihoods (i.e. $-ln(L)$ in the formula above), we can calculate the AICs and compare them.

```{r}
AIC_simple = 6 + 2*fit_negexpMLE$value

AIC_simple

AIC_full = 6 + 2*fit_negexpMLES$value

AIC_full
```

<br>

When comparing models, the lower the AIC the better, and in general a difference in AIC of 3 or more is analagous to the models being significantly different at an $\alpha$ of $p < 0.05$.

```{r}
AIC_simple - AIC_full
```

<br>

Q5: Is there a preferred model and if so, which one?

>*Answer 5:* There is a preferred model - model 2. This is because it has a lower AIC value and the difference in their AIC values is larger than 3 meaning that model 2 is significantly better than model 1.

<br>

The nice thing about AIC is that the models you compare do not have to be nested like they do for ANOVA, as long as the data are the same. There are a few other constraints however... 

Here are the AIC scores for our pair of NLS models:

```{r}
AIC(fit_negexp, fit_negexpS)
```

<br>

You'll notice that these are completely different to the AICs for the MLE models...

Q6: Why is it not okay to compare the AIC of these NLS models with the AIC of the MLE models? Hint: type `?AIC` into the R console and do some reading.

>*Answer 6:*This is because a model which is inherently better/ is the better pick in reality can have the lowest AIC value as the likelihoods of the models have not been maximised. 

<br>
